---
title: "Exploratory Data Analysis"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
source("explore.R")
```

This section explores key patterns and relationships in the dataset, focusing on how different weather conditions and time-of-day factors relate to Bike Share usage in Toronto. Visualizations are used to reveal distributions, trends, and correlations.

## Feature Distributions {.tabset}

### Bike Trips

```{r, echo=FALSE, warning=FALSE, style="display: flex; justify-content: center;"}
plot_total_trips
```

### Temperature

```{r, echo=FALSE, warning=FALSE, style="display: flex; justify-content: center;"}
plot_temp
```

### Pressure

```{r, echo=FALSE, warning=FALSE, style="display: flex; justify-content: center;"}
plot_pressure
```

### Humidity

```{r, echo=FALSE, warning=FALSE, style="display: flex; justify-content: center;"}
plot_humidity
```

### Wind Speed

```{r, echo=FALSE, warning=FALSE, style="display: flex; justify-content: center;"}
plot_wind
```

### Cloudiness

```{r, echo=FALSE, warning=FALSE, style="display: flex; justify-content: center;"}
plot_cloud
```

### Weather Conditions
```{r, echo=FALSE, warning=FALSE, style="display: flex; justify-content: center;"}
plot_weather_pie
```


## Correlation and Scatterplots {.tabset}

### Correlation

```{r, echo=FALSE, message=FALSE, warning=FALSE, style="display: flex; justify-content: center;"}
plot_total_corr
```

### Temperature

```{r, echo=FALSE, message=FALSE, warning=FALSE, style="display: flex; justify-content: center;"}
plot_temp_line
```

### Pressure

```{r, echo=FALSE, message=FALSE, warning=FALSE, style="display: flex; justify-content: center;"}
plot_pressure_line
```

### Humidity

```{r, echo=FALSE, message=FALSE, warning=FALSE, style="display: flex; justify-content: center;"}
plot_humidity_line
```

### Wind Speed

```{r, echo=FALSE, message=FALSE, warning=FALSE, style="display: flex; justify-content: center;"}
plot_wind_line
```

### Cloudiness

```{r, echo=FALSE, message=FALSE, warning=FALSE, style="display: flex; justify-content: center;"}
plot_cloudiness_line
```

## Time & Weather Effects {.tabset}

### Time-of-Day

```{r, echo=FALSE, message=FALSE, warning=FALSE, style="display: flex; justify-content: center;"}
plot_hourly_box
```

### Weather Condition

```{r, echo=FALSE, message=FALSE, warning=FALSE, style="display: flex; justify-content: center;"}
plot_weather_box
```
